---
title: "betaMix Network Analysis: `r params$dataset_name`"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    fig_height: 4
    fig_width: 6
    latex_engine: pdflatex
params:
  dataset_name: "Dataset"
  M: NULL
  res: NULL
  adjMat: NULL
  graphComp: NULL
  col_stats: NULL
  rep_cluster: 1
  signed: FALSE
  nodecol: "steelblue"
  edgecols: "grey55"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(betaMix)
```

---

# Data Overview

```{r data-dims}
N  <- nrow(params$M)
P  <- ncol(params$M)
cn <- colnames(params$M)
rn <- rownames(params$M)
```

The dataset **`r params$dataset_name`** contains **`r N`** samples (rows) and
**`r P`** variables (columns).

```{r col-names, results='asis'}
if (!is.null(cn) && P <= 40) {
  cat("\n**Variables:**", paste(cn, collapse = ", "), "\n")
} else if (P > 40) {
  cat(sprintf("\n*Variable names not shown (P = %d > 40); see summary table below.*\n", P))
}
if (!is.null(rn)) {
  cat(sprintf("\nSample identifiers are present (e.g., %s).\n",
              paste(head(rn, 3), collapse = ", ")))
}
```

## Summary Statistics

```{r col-stats, results='asis'}
cs <- params$col_stats
if (P <= 30) {
  knitr::kable(round(cs, 4),
               caption = "Per-variable summary statistics.")
} else {
  # Summarise the distribution of per-column statistics
  six_stat <- function(x) c(min(x), quantile(x, 0.25), median(x),
                              mean(x), quantile(x, 0.75), max(x))
  stat_summary <- data.frame(
    Quantity       = c("Minimum", "1st quartile", "Median", "Mean",
                       "3rd quartile", "Maximum"),
    `Column means` = round(six_stat(cs$Mean), 4),
    `Column SDs`   = round(six_stat(cs$SD),   4),
    check.names    = FALSE,
    row.names      = NULL
  )
  knitr::kable(stat_summary,
               caption = paste0("Distribution of per-column statistics across all ",
                                P, " variables."))
}
```

---

# betaMix Model

betaMix models the distribution of pairwise association statistics
$z_{ij} = 1 - r_{ij}^2 \in (0,1)$ as a mixture of two beta distributions.
Under the null (no association) $z_{ij} \sim \text{Beta}(\hat{\eta}, 0.5)$;
correlated pairs follow $\text{Beta}(\hat{a}, \hat{b})$ on the rescaled support
$[0, b_{\max}]$.

## Estimated Parameters

```{r model-params, results='asis'}
res <- params$res
param_tab <- data.frame(
  Parameter = c(
    "Sample size ($N$)",
    "Number of variables ($P$)",
    "Null proportion ($\\hat{p}_0$)",
    "Non-null shape: $\\hat{a}$",
    "Non-null shape: $\\hat{b}$",
    "Effective sample size ($\\hat{\\eta}$)",
    "Non-null support upper bound ($b_{\\max}$)",
    "Posterior probability threshold",
    "Edges detected",
    "Correlation method"
  ),
  Value = c(
    res$N,
    res$nodes,
    round(res$p0, 4),
    round(res$ahat, 3),
    round(res$bhat, 3),
    round(res$etahat, 3),
    res$bmax,
    formatC(res$ppthr, format = "e", digits = 3),
    prettyNum(res$edges, big.mark = ","),
    res$method
  )
)
knitr::kable(param_tab, col.names = c("Parameter", "Value"),
             caption = "Estimated betaMix model parameters.")
```

## Edge Detection Threshold

A variable pair $(i, j)$ is declared an **edge** when its association statistic
$z_{ij} = 1 - r_{ij}^2$ falls below the posterior probability threshold:

$$z_{ij} < \hat{\tau} = `r formatC(res$ppthr, format = "e", digits = 3)`$$

This is equivalent to requiring $|r_{ij}| >
`r round(sqrt(1 - res$ppthr), 4)`$.
The threshold $\hat{\tau}$ is chosen to control the posterior probability that
a declared edge is a false positive at the level $\text{ppr} = `r res$ppthr`$,
subject to $\alpha \leq `r formatC(res$ppthr, format = "e", digits = 3)`$ (see
`?betaMix` for details).

## Model Fit Assessment

The 2×2 diagnostic panel below assesses how well the fitted mixture describes
the data, followed by a printed verdict.

- **Top-left**: fitted mixture histogram. The green dashed curve is the null
  component $\hat{p}_0 \cdot \text{Beta}(\hat{\eta}, 0.5)$, the red curve is
  the non-null component, the blue curve is the mixture, and the orange region
  marks pairs below the threshold (detected edges).
- **Top-right**: Q-Q plot for *pure-null* observations ($m_0 \geq 0.9$),
  compared against $\text{Beta}(\hat{\eta}, 0.5)$ **conditional on** the
  sub-range spanned by those observations.  By restricting to the upper tail
  of the null distribution (small correlations, $z_{ij}$ near 1), where the
  posterior probability of being null is high, we avoid the overlap region
  with the non-null component and assess the null fit where it matters for
  inference.  Points in blue indicate a good conditional fit (KS $D \leq
  0.05$); red indicates a notable departure.  Falls back to $m_0 \geq 0.5$
  if fewer than 10 pure-null observations are available.
- **Bottom-left**: Q-Q plot for *pure-non-null* observations ($m_0 \leq 0.1$,
  scaled by $b_{\max}$), compared against $\text{Beta}(\hat{a}, \hat{b})$
  **conditional on** the sub-range of the pure-non-null subset.  Restricting
  to the lower tail of the non-null distribution (high correlations, small
  $z_{ij}/b_{\max}$) removes the right-side overlap where the null component
  dominates.  Because the non-null Beta is a convenience model, deviations
  in the overlap region do not affect inference; the conditional Q-Q focuses
  on the region that does.  Falls back to $m_0 < 0.5$ if fewer than 10
  pure-non-null observations are available.
- **Bottom-right**: model-based ROC curve sweeping $\tau$ from 0 to 1 using
  the fitted Beta distributions. The orange point marks the current threshold
  $\hat{\tau}$; the red point marks the Youden-index optimum
  $\tau^* = \arg\max_\tau [\text{TPR}(\tau) - \text{FPR}(\tau)]$.
  The grey dotted vertical line shows the FPR at $\tau = b_{\max}$.

The KS $D$ statistic measures practical deviation; raw p-values are not used
as the threshold because with tens of thousands of pairs even $D \approx 0.01$
yields $p \approx 0$.

```{r assess-fit, fig.height=7, fig.width=7, fig.cap="Goodness-of-fit diagnostics: fitted mixture (top-left), null Q-Q (top-right), non-null Q-Q (bottom-left), model-based ROC curve (bottom-right). Blue Q-Q points indicate good fit (KS D <= 0.05); red indicates notable departure."}
assessFit(params$res)
```

---

# Network Analysis

```{r network-summary, results='asis'}
res    <- params$res
gc     <- params$graphComp
nc     <- max(gc$clustNo)
n_uncl <- sum(gc$clustNo == 0)
cat(sprintf(
  "The estimated network has **%s nodes** and **%s edges**",
  prettyNum(res$nodes, big.mark = ","),
  prettyNum(res$edges, big.mark = ",")))
cat(sprintf(
  " (out of %s possible pairs).\n",
  prettyNum(choose(res$nodes, 2), big.mark = ",")))
cat(sprintf(
  "Graph decomposition (minimum cluster size = 2) yields **%d cluster(s)** and **%d unclustered node(s)**.\n",
  nc, n_uncl))
```

## Cluster Summary

```{r cluster-table, results='asis'}
invisible(capture.output(clusterTab <- summarizeClusters(params$graphComp)))
if (!is.null(clusterTab) && nrow(clusterTab) > 0) {
  tab_display <- clusterTab
  pct_cols <- grep("pctInClst", colnames(tab_display))
  tab_display[, pct_cols] <- round(tab_display[, pct_cols], 3)

  deg_cols <- c(1, 2, 3, 4, 5, 6, 7)
  pct_cols_idx <- c(1, 8, 9, 10, 11, 12)

  print(knitr::kable(tab_display[, deg_cols],
               col.names = c("Cluster", "Nodes",
                             "Min", "Q1", "Median", "Q3", "Max"),
               booktabs  = TRUE,
               row.names = FALSE,
               caption   = "Cluster degree summary (five-number summary of node degree within each cluster)."))
  cat("\n")
  print(knitr::kable(tab_display[, pct_cols_idx],
               col.names = c("Cluster",
                             "Min", "Q1", "Median", "Q3", "Max"),
               booktabs  = TRUE,
               row.names = FALSE,
               caption   = "Proportion of within-cluster edges (five-number summary of the fraction of each node's edges that lie within its cluster).",
               digits    = 3))
} else {
  cat("*No clusters detected (all nodes are unclustered).*\n")
}
```

## Degree vs. Clustering Coefficient

Each point represents one node. The x-axis is node degree; the y-axis is the
product degree × CC, where CC is the clustering coefficient (fraction of a
node's neighbour pairs that are themselves connected). The green diagonal
($y = x$) marks nodes with CC = 1 (all neighbours are mutually connected).

```{r plot-degcc, fig.height=4, fig.width=5, fig.cap="Node degree vs. degree × clustering coefficient. The green line is the CC = 1 reference."}
plotDegCC(params$res, params$graphComp)
```

## Adjacency Matrix

Nodes are ordered by cluster membership. A black pixel at position $(i, j)$
indicates a detected edge between nodes $i$ and $j$.

```{r plot-bitmap, fig.height=4, fig.width=4, fig.cap="Adjacency matrix ordered by cluster. Black pixels are edges."}
plotBitmapCC(params$adjMat, params$graphComp, orderByCluster = TRUE)
```

## Representative Cluster Network

```{r cluster-info}
rc        <- params$rep_cluster
gc        <- params$graphComp
n_in_rc   <- if (!is.na(rc)) sum(gc$clustNo == rc) else 0
```

```{r plot-cluster, fig.height=5, fig.width=5, fig.cap=paste0("Network of cluster ", rc, " (", n_in_rc, " nodes). Node size is proportional to degree; opacity reflects the fraction of a node's edges that lie within the cluster.")}
if (!is.na(params$rep_cluster)) {
  plotCluster(
    params$adjMat,
    params$rep_cluster,
    params$graphComp,
    labels   = TRUE,
    nodecol  = params$nodecol,
    edgecols = params$edgecols
  )
} else {
  plot.new()
  text(0.5, 0.5, "No clusters detected.", cex = 1.2)
}
```

---

*Report generated by the **betaMix** R package.*
*`r R.version.string`, betaMix `r packageVersion("betaMix")`.*
